I don't know, are you sure we want to do this?\\ Sea level adaptation decisions under uncertainty

### 2016-12-12 submission/paper.tex
Added material to introduction

\section{Introduction}\label{sec:intro}
The potential impact of climate change on local sea level, yielding effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats, requires city planners to make decisions in the presence of substantial uncertainty.

The Norwegian city of Bergen is the capital of Hordaland County. The city center is located on Byfjorden, and is surrounded by mountains. It has the largest port in Norway, both in terms of freight and passengers. The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UNESCO World Heritage site (\url{http://whc.unesco.org/en/list/59}). Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become a major problem in other parts of Bergen as well \citep{bergenreport}.

Esbjerg, on the west coast of Jutland,  has the second largest harbour in Denmark. It can be subjected to substantial storm surges, which, in conjunction with heavy rainfall, can flood substantial parts of the town.

In this paper we illustrate the uncertainty of sea level projections, and combine this with a simplified decision problem where planners want to know how early they should implement costly adaptation measures.

Section \ref{sealevelproj} describes our approach to projecting sea level. In section \ref{decision_tools} we describe the type of decision problem that we are going to attack. We apply these tools to  sea level projections for Bergen, Norway, and for Esbjerg, Denmark 
in section \ref{cases}.  In the final section \ref{unc}  we demonstrate the consequences of ignoring the uncertainty in the projections.

>>>>

\section{Introduction}\label{sec:intro}
Long-term planning and decision-making regarding societal infrastructure along coastal areas must account for a changing climate. Especially as the potential impact of climate change on local sea level can yield numerous effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats (add REF). Considerable challenges continue to exist in understanding  the impacts of a combination of sea level rise and storm surges and the implementation of potential adaptation options to reduce these impacts. Severe inherent uncertainties persist along all parts of the processing chain from climate projections to adaptation assessments, and it is critical that decision-making appropriately accounts for this.

As adaptation decision-making is an ongoing process of weighing and choosing which measures should be taken at which moment in time \citep{Hallegatte&2012} adaptive planning methods need to support decisions in the short term, while considering long-term developments. Challenges of adaptation decision-making under uncertainty relate to the incorporation of spatial, inter-temporal and flexibility aspects of adaptation priorities \citep{FankhauserSoare2013}, and the linkage with specific characteristics of sectors and contexts \citep{BisaroSwartHinkel2016, HinkelBisaro2016}. Several economic decision support tools and methods exist for adaptation assessment under uncertainty \citep[e.g.][]{Chambwera&2014, WilbyDessai2010, WalkerHaasnootKwakkel2013}. However, \citet{Watkiss&2015} conclude that these tools are very resource intensive and complex in the context of long-term adaptation investment decisions and call for the development of ``light touch'' approaches to better support local adaptation making.

In this paper we combine the uncertainty of sea level projections with a simplified decision framework under uncertainty where planners want to know how early they should implement costly adaptation measures.  We discuss the advantages of using a fully probabilistic approach in a light touch decision framework that accounts for uncertainty both in climate projections and damage assessments, where information on uncertainty is turned into a driver of change rather than a barrier.

We illustrate the sea level rise projections and timing of adaptation measures for two case studies in Northern Europe, namely the city of Bergen in Norway and Esbjerg in Denmark. The Norwegian city of Bergen is the capital of Hordaland County. The city center is located on Byfjorden, and is surrounded by mountains. It has the largest port in Nor way, both in terms of freight and passengers. The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UN-ESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}. Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become  a major problem in other parts of Bergen as well \citep{bergenreport}. Esbjerg, on the west coast of Jutland, has the second largest harbour in Denmark. It can be subjected to substantial storm surges, which, in conjunction with heavy rainfall, can flood substantial parts of the town.

Section \ref{sealevelproj} describes our approach to projecting sea level. In section \ref{decision_tools} we describe the decision framework focused on the timing of proactive adaptation decision. We apply this framework to sea level projections and adaptation for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}. In the final section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections.
